---
title: "the art of soccer economics"
output:
  html_document: default
  
---

<br>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

packages <- c("ggplot2")
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

rm(list=ls())
season1920 <- read.csv(file="/Users/SimonThielen/Desktop/R/Daten/2021_02_07_Football/2010s/2019_20/de1.csv")
```
    
  
## introduction

For my first project I turn to a familiar topic - soccer. Since my childhood I am passionate about soccer. I played in several teams, trained some kids, tracked nearly every game in the highest european divisions and discussed a lot with friends about games and transfers. So I am pretty sure to know everything about this sport. But do my knowledge hold against the raw data?  
In this project I want to test some general hypothesis and check I the data confirm or contradict my view on soccer. On the following pages I will first introduce the data, then state my hypothesis and finally check if the data support my hypothesis.  
  
  
## show me the data 
I found the datasets on GitHub^[https://github.com/jokecamp/FootballData/tree/master/Germany]. A User provides the results of the german Bundesliga, with matchday number, matchday date, teams and result. I add a factor variable with levels 1 to 3 that states the result as 'Draw', 'HomeVictory' and 'AwayVictory', as well as two integer variables for the goals. At least I add another dummy variable that states if the victory of a team was narrow, means a victory by one goal difference. The summary statistics of the datasets looks as follows. Sadly we could not get much valueable information out of this first summary. Therefore, we now have to perfom some statistics.

```{r summary}

season1920$Team.1Goals <- as.integer(sub("-.*", "", season1920$FT)) 
season1920$Team.2Goals <- as.integer(sub("*.-", "", season1920$FT)) 
season1920$result <- factor(ifelse(season1920$Team.1Goals==season1920$Team.2Goals,0,
       ifelse(season1920$Team.1Goals>season1920$Team.2Goals,1,2)),levels=c(0,1,2),
       labels=c("Draw","HomeVictory","AwayVictory"))
season1920$narrowVict <-factor(ifelse(abs(season1920$Team.1Goals-season1920$Team.2Goals)==1,
                                      1,0))
summary(season1920)
```


## manipulate the data
For all that can't remember preciously about the season I reproduce the final table. Using the aggregate function and ordering the result descending gives the following ranking. Like every year the Bayern won the league and qualified for the champions league. Dortmund, Liepzig and Gladbach also qualified for the champions league. Paderborn and Düsseldorf had to go to the 2. Bundesliga and Bremen keep the class in relegation. There is one small mistake in the table. Can you spot and correct it?

```{r finalTable}
season1920$Team.1Points <- ifelse(season1920$Team.1Goals>season1920$Team.2Goals,3,
                                  ifelse(season1920$Team.1Goals==season1920$Team.2Goals,
                                         1,0))
season1920$Team.2Points <- ifelse(season1920$Team.1Goals<season1920$Team.2Goals,3,
                                  ifelse(season1920$Team.1Goals==season1920$Team.2Goals,
                                         1,0))
finalTableHome <- aggregate(x=season1920[, c(6,7,10)], by=list(season1920$Team.1),
                            FUN=sum)
colnames(finalTableHome) <-c("Team", "GoalsScored", "GoalsAgainst", "Points")
finalTableAway <- aggregate(x=season1920[, c(6,7,11)], by=list(season1920$Team.2),
                            FUN=sum)
colnames(finalTableAway) <-c("Team", "GoalsAgainst", "GoalsScored", "Points")
finalTableAway <- finalTableAway[ ,c(1, 3, 2, 4)] 
finalTable <- rbind(finalTableHome, finalTableAway)
finalTable <- aggregate(x=finalTable[, c(2,3,4)], by=list(finalTable$Team),
                        FUN=sum)
colnames(finalTable) <-c("Team", "GoalsScored", "GoalsAgainst", "Points")
finalTable <- finalTable[order(finalTable$Points, decreasing = TRUE),]
rownames(finalTable) <- 1:18
print(finalTable)
```

Now we want to know if their are teams that perform significantly better at home than on the opponents pitch. Therefore, we compare gained points from the home and away table. It seems that teams on the upper half of the table have highly different results at home and away why teams on the margins make nearly the same amount of points at home and away. Only Bremen seems to be an exception from this finding. The question is if this result is by chance or might there be an explanation for this finding. I will load more data, namely the seasons from 2016 until 2018 and look for a similar effect.

```{r finalTableAwayHome}
colnames(finalTableHome) <-c("Team", "GoalsScored", "GoalsAgainst", "PointsHome")
colnames(finalTableAway) <-c("Team", "GoalsScored", "GoalsAgainst", "PointsAway")
finalTable <- merge(finalTable, finalTableHome[,c(1,4)], by="Team")
finalTable <- merge(finalTable, finalTableAway[,c(1,4)], by="Team")
finalTable <- finalTable[order(finalTable$Points, decreasing = TRUE),]
finalTable$Diff <- finalTable$PointsHome - finalTable$PointsAway
rownames(finalTable) <- 1:18
print(finalTable[ ,c(1,4,5,6,7)])
```


## add more data

The new seasons are loaded from the same gitHub source and contain additional three seasons. The data manipulation is done in the same way as before and the results are stored in one data frame. A new character varible season is introduced to differentiate the seasons.

```{r additionalSeasons}

finalTable1920 <- finalTable
finalTable1920$season <- "19/20"
rm(finalTable)

for (i in 16:18) {
  season <- read.csv(file=paste0("/Users/SimonThielen/Desktop/R/Daten/2021_02_07_Football/2010s/20",
                                                 i,"_",i+1,"/de1.csv"))
  
  season$Team.1Goals <- as.integer(sub("-.*", "", season$FT)) 
  season$Team.2Goals <- as.integer(sub("*.-", "", season$FT)) 
  season$result <- factor(ifelse(season$Team.1Goals==season$Team.2Goals,0,
       ifelse(season$Team.1Goals>season$Team.2Goals,1,2)),levels=c(0,1,2),
       labels=c("Draw","HomeVictory","AwayVictory"))
  season$narrowVict <-factor(ifelse(abs(season$Team.1Goals-season$Team.2Goals)==1,
                                      1,0))

  season$Team.1Points <- ifelse(season$Team.1Goals>season$Team.2Goals,3,
                                  ifelse(season$Team.1Goals==season$Team.2Goals,
                                         1,0))
  season$Team.2Points <- ifelse(season$Team.1Goals<season$Team.2Goals,3,
                                  ifelse(season$Team.1Goals==season$Team.2Goals,
                                         1,0))
  finalTableHome <- aggregate(x=season[, c(6,7,10)], by=list(season$Team.1),
                            FUN=sum)
  colnames(finalTableHome) <-c("Team", "GoalsScored", "GoalsAgainst", "Points")
  finalTableAway <- aggregate(x=season[, c(6,7,11)], by=list(season$Team.2),
                            FUN=sum)
  colnames(finalTableAway) <-c("Team", "GoalsAgainst", "GoalsScored", "Points")
  finalTableAway <- finalTableAway[ ,c(1, 3, 2, 4)] 
  finalTable <- rbind(finalTableHome, finalTableAway)
  finalTable <- aggregate(x=finalTable[, c(2,3,4)], by=list(finalTable$Team),
                        FUN=sum)
  colnames(finalTable) <-c("Team", "GoalsScored", "GoalsAgainst", "Points")
  finalTable <- finalTable[order(finalTable$Points, decreasing = TRUE),]
  rownames(finalTable) <- 1:18
  colnames(finalTableHome) <-c("Team", "GoalsScored", "GoalsAgainst", "PointsHome")
  colnames(finalTableAway) <-c("Team", "GoalsScored", "GoalsAgainst", "PointsAway")
  finalTable <- merge(finalTable, finalTableHome[,c(1,4)], by="Team")
  finalTable <- merge(finalTable, finalTableAway[,c(1,4)], by="Team")
  finalTable <- finalTable[order(finalTable$Points, decreasing = TRUE),]
  finalTable$Diff <- finalTable$PointsHome - finalTable$PointsAway
  rownames(finalTable) <- 1:18
  finalTable$season <- paste0(i,"/",i+1)
  assign(paste0("finalTable",i,i+1),finalTable) 
  
  rm(season, finalTable, finalTableHome, finalTableAway)
}

finalTable <- rbind(finalTable1617, finalTable1718, finalTable1819, finalTable1920)
finalTable$final <-1:18
```
  
  
## set up some hypothesis
Now that the data is loaded and transformed I can state some hypothesis to test on the data. The first hypothesis I already mentioned in the previous section is about the correlation between the rank of a team in the final table and it's different outcomes at home and away. The hypothesis is therefore:  
  
**H0: The rank in the table is independent of the difference between the outcome at home and away**    
**H1: It is not**  
  
The second Hypothesis I want to adress is if there is an effect is it stronger on the midfield of the table than on the edges?  
  
**H0: The differents in home and away results is independet of the in the final table the team end up**  
**H1: It is not**  
  
[//]: # And the last hypothesis would be if there is any correlation between a larger gap of home and away points and tight wins or losses  
[//]: # **H0: The teams amount of tight wins or losses is independet of his gap between home and away points**  
[//]: # **H1: It is not**  
  
    
## descriptive statistics
Before I do some specific statistics on the hypothesis, I will have a quick look over the general data. The summary table shows the 5-point-summary of the variable Diff. with a max of 25, which means 25 point more at home than away and a min of -13, which mean 13 points more away than at home. The median at 7 is slidly lower than the mean with 7.25 which suggest a skewness to the right, which can be observed on the histogram plot. The standard deviation is 7.66 by a rage of 38, with can be interpreted as a moderate spread of the data. The boxplots show that there are hughe differences between the teams over the different seasons, but there are also some teams that always have differences about the same amount and this is mostly around zero. Sometimes there are teams that have seem to have huge and constant differences, like Darmstadt or Paderborn, but this is due to the effect that they only stayed in the first devision for one season. This can be better observed in the time series plot where these clubs only have one dot. Also the graphs in this plot cross each other very often which is a sign that the differences are not constant over time. In the next section we now dig deeper into the hypotheses testing.
```{r descriptiveStatistics1}
summary(finalTable)
sd(finalTable$Diff)

hist(finalTable$Diff, main="Distribution Differences Home/Away") 

plot(as.factor(finalTable$Team), finalTable$Diff, las=3)

library("ggplot2")
ggplot(data = finalTable, aes(x = factor(season), y = Diff, color = Team)) +       
  geom_line(aes(group = Team)) + geom_point()

```


## hypothesis testing

To test the hypothesis I first conduct a robust variance analysis with the kruskal test. With a p-value of 0.2996 it is not possible to reject the Nullhypothesis and we can therefore not assume an effect that is different from zero. An logit regression also failed to find significant effects. The problem might be the small amount of datapoints I have for each team. Therefore I intersect the final tables in 4 sections. The first section includes position 1 to 4 (these are the authorize to play champions league), the section sectionen with position 5 to 9, the third with positions 10 to 14 and the last with 15 to 18. Now I try the same analysis again and look for significant results. This is equivalent with our second hypothesis, that the area a team ends in is important. And again I do not fina a significant effect.
```{r hypothesisTesting1}
kruskal.test(finalTable$Team, finalTable$Diff) 
log1 <- glm(Diff ~ Team, data=finalTable)
summary(log1)
```

```{r hypothesisTesting2}
finalTable$section <- ifelse(finalTable$final<=4, 1,
                             ifelse(finalTable$final<=9,2,
                                 ifelse(finalTable$final<=14,3,4
                                        )))
log2 <- glm(Diff ~ section, data=finalTable)
summary(log2)
```


## conclusion
In this project I tried to find support signs for the hypothesis that the difference beetwen the performance of a soccer team at home and away has an effect on their position at the end of a season. With the presented data it was not possible to identify such effects. Next steps could be extending the data by more seasons or trying other table sections, for example splitting the table only in two sides or in a middle and edge part.

